Galaxy Clustering Constraints on the Galaxy Size--Halo Connection

### 2017-09-02 paper/galsize_paper.tex
Clarified notation on rhalf

%%%%%%%%  galaxy properties  %%%%%%%%
\newcommand{\rhalf}{R_{1/2}}
\newcommand{\rhalfdisk}{R_{1/2}^{\rm disk}}
\newcommand{\rhalfbulge}{R_{1/2}^{\rm bulge}}
\newcommand{\adisk}{A_{\rm disk}}
\newcommand{\abulge}{A_{\rm bulge}}
\newcommand{\alphadisk}{\alpha_{\rm disk}}
\newcommand{\alphabulge}{\alpha_{\rm bulge}}
\newcommand{\sigmarhalf}{\sigma_{\rm R_{1/2}}}
\newcommand{\rvir}{R_{\rm vir}}
\newcommand{\bt}{{\rm B/T}}
\newcommand{\mstar}{M_{\ast}}
\newcommand{\ssfr}{{\rm sSFR}}
\newcommand{\sfr}{{\rm SFR}}

@@

%%%%%%%%  units  %%%%%%%%
\newcommand{\kpc}{{\rm kpc}}
\newcommand{\mpc}{{\rm Mpc}}
\newcommand{\msun}{M_\odot}

@@

\begin{abstract}
We derive empirical modeling constraints on the connection between dark matter halos and the half-mass radius $\rhalf$ of galaxy bulges and disks. We show that both $\rhalfdisk$ and $\rhalfbulge$ are well-described by power law scaling relations with halo virial radius, $\rhalf=A\rvir^{\alpha}.$ Novel to this work, we use new SDSS measurements of the $\rhalf-$dependence of galaxy clustering to constrain the model parameters, $A_{\rm bulge}, A_{\rm disk}, \alpha_{\rm bulge},\alpha_{\rm disk},$ and log-normal scatter $\sigmarhalf.$ Even when only coarsely tuning these parameters to the observed one-point functions $\langle\rhalfdisk|\mstar^{\rm disk}\rangle$ and $\langle\rhalfbulge|\mstar^{\rm bulge}\rangle,$ our model accurately predicts the observed two-point clustering on small- and large-scales. This success non-trivial, as we show that galaxy clustering is highly sensitive to the physics that shapes satellite galaxy profiles. We find no evidence for the commonly assumed relation between halo spin $\halospin$ and $\rhalfdisk,$ and show that this assumption cannot be meaningfully constrained with either the clustering or lensing of  $L_{\ast}$ galaxies. Our results provide simple boundary conditions for more complex and fine-grained models of galaxy size. We make our python code publicly available to support cosmological surveys that require realistic synthetic galaxy populations.
\end{abstract}

@@

We additionally use the catalog of SDSS galaxy profile decompositions provided by \citet{meert_etal15}. In the version of this catalog that we use, two-dimensional $r-$band profiles were fit with a two-component de Vaucouleurs + exponential profile. We cross-match our DR10 sample against the \citet{meert_etal15} catalog values of $\rhalf,$ the half-light radius of the full profile, and $\bt,$ the $r-$band flux ratio of the de Vaucouleurs component to the total $r-$band flux. This cross-matching gives us a volume-limited galaxy sample with values of $\mstar,$ $\ssfr,$ $\rhalf,$ and $\bt.$ Together with the present work, we make publicly available this cross-matched catalog, as well as code to generate the catalog from the original data files and load it into a python interpreter.

We calculate two-point clustering $\wproj$ of our SDSS galaxy sample using line-of-sight projection of $\pi_{\rm max}=20\mpc$ using the {\tt correl} program in {\tt UniverseMachine}. Our results in \S~\ref{sec:results} will give special focus on the dependence of $\wproj$ upon composite galaxy size $\rhalf.$ We will quantify this dependence in terms of {\em clustering ratios} of ``large" vs. ``small" galaxies, defined according to whether composite galaxy size is above or below $\langle\rhalf\vert\mstar\rangle,$ computed as the median of a sliding stellar mass window with a width of $N_{\rm gal}=1000.$

@@

For central galaxies residing in host halos, we use the present-day virial radius; for satellite galaxies we use the virial radius at the time of accretion, using physical units of $\kpc.$ We model the composite size of model galaxies as 
\beq
\rhalf = (\bt)\rhalfbulge + (1-\bt)\rhalfdisk.
\eeq

@@

We reiterate an important distinction drawn in the formulation of our forward modeling approach. In the model described above, $\rhalfbulge$ and $\rhalfdisk$ are forward modeling quantities that we aim to constrain through SDSS observations of {\em composite} galaxy size $\rhalf.$ In particular, both model and real galaxies will be categorized as ``small" or ``large`` according to whether $\rhalf$ is above or below the median size for a galaxy with total stellar mass $\mstar.$

@@

In Figure \ref{fig:scatter_plot} we show the scaling of composite galaxy size $\rhalf$ with $\mstar,$ with separate panels dedicated to subsamples of galaxies divided according to value of $\bt.$ We refer to galaxies with $\bt<0.25$ as ``disk-dominated", $\bt>=0.75$ as ``bulge-dominated", and $0.25<=\bt<0.75$ as ``mixed morphology".

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/size_vs_stellar_mass_multipanel_bt_decomposition.pdf}
\caption{
{\bf One-point data used to fit the fiducial model.}
Scattered points show the $\rhalf-\mstar$ relation for SDSS galaxies as measured in \citet{meert_etal15}. Bulge-dominated galaxies are defined in terms of the bulge-to-total stellar mass ratio $\bt>=0.75,$ disk-dominated galaxies $\bt<0.25,$ mixed morphology as $0.25 < \bt < 0.75.$ The black curve in each panel shows the $\rhalf-\mstar$ relation implied by our fiducial model, in which $\rhalf=A\rvir^\alpha,$ with $\adisk=0.014=7\abulge, \alphadisk=1, \alphabulge=5/4,$ and uncorrelated log-normal scatter of $0.2$ dex about these relations. 
}
\label{fig:scatter_plot}
\end{figure}
%-----------------------------------------------------------------------------------------------------

@@

Comparison of the top panels of Fig.~\ref{fig:scatter_plot} show that this linearization is only approximate; evidently, the composite size $\rhalf$ of bulge-dominated galaxies has a steeper scaling relation with $\mstar$ relative to disk-dominated galaxies. This difference motivates our simple extension of the \citet{kravtsov13} model, in which composite size $\rhalf$ of disk- and bulge-dominated galaxies scale according to separate power laws.

@@

To achieve this, we compute projected clustering separately for $\mstar-$complete samples of galaxies defined by simultaneous cuts on $\bt$ and $\rhalf,$ described as follows. First, we construct subsamples of disk-dominated and bulge-dominated galaxies using cuts of $\bt<0.25$ and $\bt>0.75,$ respectively, as in \S~\ref{subsubsec:scaling_relation} above. Separately for each $\bt-$selected subsample, we calculate $\langle\rhalf\vert\mstar;\bt\rangle,$ the median composite size of the sample as a function of total galaxy stellar mass. We split each $\bt-$selected sample in half according to whether the galaxy has a composite size $\rhalf$ that is above or below this median value. For any $\mstar-$complete threshold, this gives four subsamples of large and small, disk- and bulge-dominated galaxies. For each $\bt-$selected sample, we calculate the difference $\wproj^{\rm large}-\wproj^{\rm small},$ scaled by $\wproj^{\rm all},$ the clustering of the $\bt-$selected sample prior to splitting on $\rhalf,$ and refer to this quantity as {\em the $\rhalf$ clustering ratio}.

@@

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/random_bt_centrals_vs_satellites.pdf}
\caption{
{\bf Morphologies of centrals and satellites of the same halo mass.} We show the PDF of the bulge-to-total stellar mass ratio $\bt$ of central and satellite galaxies of the same halo mass $\mhalo=\mpeak.$ In the model, the PDF of $\bt$ is determined by the stellar mass $\mstar$ and specific star-formation rate $\ssfr$ statistically determine, with no residual dependence on the cosmic web. Satellite galaxies in the model are ``bulgier" than centrals of the same halo mass because satellites are more quiescent than centrals of the same stellar mass. Thus our baseline model ansatz is that the morphology-density relation is purely derived from the color-density relation. Since composite size $\rhalf$ in the model is given by $\rhalf=(\bt)\rhalfbulge + (1-\bt)\rhalfdisk,$ this ansatz in turn implies that smaller galaxies cluster more strongly relative to larger galaxies of the same stellar mass, as seen in the measurements shown in Fig.~\ref{fig:clustering_ratio_upshot}. 
}
\label{fig:bt_censat}
\end{figure}
%-----------------------------------------------------------------------------------------------------

>>>>

%%%%%%%%  galaxy properties  %%%%%%%%
\newcommand{\rhalf}{R_{1/2}}
\newcommand{\rhalftot}{R^{\rm tot}_{1/2}}
\newcommand{\rhalfdisk}{R_{1/2}^{\rm disk}}
\newcommand{\rhalfbulge}{R_{1/2}^{\rm bulge}}
\newcommand{\adisk}{A_{\rm disk}}
\newcommand{\abulge}{A_{\rm bulge}}
\newcommand{\alphadisk}{\alpha_{\rm disk}}
\newcommand{\alphabulge}{\alpha_{\rm bulge}}
\newcommand{\sigmarhalf}{\sigma_{\rm R_{1/2}}}
\newcommand{\rvir}{R_{\rm vir}}
\newcommand{\bt}{{\rm B/T}}
\newcommand{\mstar}{M_{\ast}}
\newcommand{\mstartot}{M^{\rm tot}_{\ast}}
\newcommand{\ssfr}{{\rm sSFR}}
\newcommand{\sfr}{{\rm SFR}}

@@

%%%%%%%%  units  %%%%%%%%
\newcommand{\kpc}{{\rm kpc}}
\newcommand{\mpc}{{\rm Mpc}}
\newcommand{\msun}{M_\odot}
\newcommand{\kms}{{\rm km/s}}

@@

\begin{abstract}
We derive empirical modeling constraints on the connection between dark matter halos and the half-mass radius $\rhalf$ of galaxy bulges and disks. We show that both $\rhalfdisk$ and $\rhalfbulge$ are well-described by power law scaling relations with halo virial radius, $\rhalf=A\rvir^{\alpha}.$ Novel to this work, we use new SDSS measurements of the $\rhalftot-$dependence of galaxy clustering to constrain the model parameters, $A_{\rm bulge}, A_{\rm disk}, \alpha_{\rm bulge},\alpha_{\rm disk},$ and log-normal scatter $\sigmarhalf.$ Even when only coarsely tuning these parameters to the observed one-point functions $\langle\rhalfdisk\vert\mstartot\rangle$ and $\langle\rhalfbulge|\mstartot\rangle,$ our model accurately predicts the observed two-point clustering on small- and large-scales. This success non-trivial, as we show that galaxy clustering is highly sensitive to the physics that shapes satellite galaxy profiles. We find no evidence for the commonly assumed relation between halo spin $\halospin$ and $\rhalfdisk,$ and show that this assumption cannot be meaningfully constrained with either the clustering or lensing of  $L_{\ast}$ galaxies. Our results provide simple boundary conditions for more complex and fine-grained models of galaxy size. We make our python code publicly available to support cosmological surveys that require realistic synthetic galaxy populations.
\end{abstract}

@@

We additionally use the catalog of SDSS galaxy profile decompositions provided by \citet{meert_etal15}. In the version of this catalog that we use, two-dimensional $r-$band profiles were fit with a two-component de Vaucouleurs + exponential profile. We cross-match our DR10 sample against the \citet{meert_etal15} catalog values of $\rhalftot,$ the half-light radius of the full (composite) profile, and $\bt,$ the $r-$band flux ratio of the de Vaucouleurs component to the total $r-$band flux. This cross-matching gives us a volume-limited galaxy sample with values of $\mstar,$ $\ssfr,$ $\rhalftot,$ and $\bt.$ Together with the present work, we make publicly available this cross-matched catalog, as well as code to generate the catalog from the original data files and load it into a python interpreter.

We calculate two-point clustering $\wproj$ of our SDSS galaxy sample using line-of-sight projection of $\pi_{\rm max}=20\mpc$ using the {\tt correl} program in {\tt UniverseMachine}. Our results in \S~\ref{sec:results} will give special focus on the dependence of $\wproj$ upon $\rhalftot.$ We will quantify this dependence in terms of {\em clustering ratios} of ``large" vs. ``small" galaxies, defined according to whether composite galaxy size is above or below $\langle\rhalftot\vert\mstartot\rangle,$ computed as the median of a sliding stellar mass window with a width of $N_{\rm gal}=1000.$

In all sections that follow, we will declutter our notation by referring to $\rhalftot$ as $\rhalf$ and $\mstartot$ as $\mstar.$ In all cases where we wish to refer to the half-light radius or stellar mass of a particular component, that component will explicitly appear in superscript. All numerical values of $\rhalf$ will be quoted in $\kpc,$ and all values of $\mstar$ and $\mhalo$ in $\msun,$ assuming $H_0=67.8~\kms\equiv100h~\kms$ the best-fit value from \citet{planck15}. To scale stellar masses to ``$h=1$ units" \citep{croton13}, our numerically quoted values for $\mstar$ should be multiplied by a factor of $h^2,$ while our halo masses and distances should be multiplied by a factor of $h.$

@@

For central galaxies residing in host halos, we use the present-day virial radius; for satellite galaxies we use the virial radius at the time of accretion, using physical units of $\kpc.$ We model the composite size of model galaxies as
\beq
\rhalftot = (\bt)\rhalfbulge + (1-\bt)\rhalfdisk.
\eeq

@@

We reiterate an important distinction drawn in the formulation of our forward modeling approach. In the model described above, $\rhalfbulge$ and $\rhalfdisk$ are forward modeling quantities that we aim to constrain through SDSS observations of {\em composite} galaxy size $\rhalftot.$ In particular, both model and real galaxies will be categorized as ``small" or ``large`` according to whether $\rhalftot$ is above or below the median size for a galaxy with total stellar mass $\mstartot.$

@@

In Figure \ref{fig:scatter_plot} we show the scaling of composite galaxy size $\rhalf$ with $\mstar,$ with separate panels dedicated to subsamples of galaxies divided according to value of $\bt.$ We refer to galaxies with $\bt<0.25$ as ``disk-dominated", $\bt>=0.75$ as ``bulge-dominated", and $0.25<=\bt<0.75$ as ``mixed morphology". We remind the reader that $\rhalf$ and $\mstar$ refer to the scale radius and stellar mass of the {\em entire} galaxy, not of an individual component.

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/size_vs_stellar_mass_multipanel_bt_decomposition.pdf}
\caption{
{\bf One-point data used to fit the fiducial model.}
Scattered points show the $\rhalf-\mstar$ relation for SDSS galaxies as measured in \citet{meert_etal15}. Bulge-dominated galaxies are defined in terms of the bulge-to-total stellar mass ratio $\bt>=0.75,$ disk-dominated galaxies $\bt<0.25,$ mixed morphology as $0.25 < \bt < 0.75.$ The black curve in each panel shows the $\rhalf-\mstar$ relation implied by our fiducial model, in which $\rhalfbulge=\abulge\rvir^{\alphabulge}$ and $\rhalfdisk=\adisk\rvir^{\alphadisk},$ with $\adisk=0.014=7\abulge, \alphadisk=1, \alphabulge=5/4,$ and uncorrelated log-normal scatter of $0.2$ dex about these relations.
}
\label{fig:scatter_plot}
\end{figure}
%-----------------------------------------------------------------------------------------------------

@@

Comparison of the top panels of Fig.~\ref{fig:scatter_plot} show that this linearization is only approximate; evidently, galaxy size $\rhalf$ of bulge-dominated systems has a steeper scaling relation with $\mstar$ relative to disk-dominated systems. This difference motivates our simple extension of the \citet{kravtsov13} model, in which $\rhalf$ of disk- and bulge-dominated galaxies scales according to separate power laws.

@@

To achieve this, we compute projected clustering separately for $\mstar-$complete samples of galaxies defined by simultaneous cuts on $\bt$ and $\rhalf,$ described as follows. First, we construct subsamples of disk-dominated and bulge-dominated galaxies using cuts of $\bt<0.25$ and $\bt>0.75,$ respectively, as in \S~\ref{subsubsec:scaling_relation} above. Separately for each $\bt-$selected subsample, we calculate $\langle\rhalf\vert\mstar;\bt\rangle,$ the median composite size of the sample as a function of total galaxy stellar mass. We split each $\bt-$selected sample in half according to whether the galaxy has a size $\rhalf$ that is above or below this median value. For any $\mstar-$complete threshold, this gives four subsamples of large and small, disk- and bulge-dominated galaxies. For each $\bt-$selected sample, we calculate the difference $\wproj^{\rm large}-\wproj^{\rm small},$ scaled by $\wproj^{\rm all},$ the clustering of the $\bt-$selected sample prior to splitting on $\rhalf,$ and refer to this quantity as {\em the $\rhalf$ clustering ratio}.

@@

%---------------------------------------------------------------------------------------------------
\begin{figure}
\centering
\includegraphics[width=8cm]{FIGS/random_bt_centrals_vs_satellites.pdf}
\caption{
{\bf Morphologies of centrals and satellites of the same halo mass.} We show the PDF of the bulge-to-total stellar mass ratio $\bt$ of central and satellite galaxies of the same halo mass $\mhalo=\mpeak.$ In the model, the PDF of $\bt$ is determined by the stellar mass $\mstar$ and specific star-formation rate $\ssfr$ statistically determine, with no residual dependence on the cosmic web. Satellite galaxies in the model are ``bulgier" than centrals of the same halo mass because satellites are more quiescent than centrals of the same stellar mass. Thus our baseline model ansatz is that the morphology-density relation is purely derived from the color-density relation. Since composite size $\rhalf$ in the model is given by $\rhalf\equiv\rhalftot=(\bt)\rhalfbulge + (1-\bt)\rhalfdisk,$ this ansatz in turn implies that smaller galaxies cluster more strongly relative to larger galaxies of the same stellar mass, as seen in the measurements shown in Fig.~\ref{fig:clustering_ratio_upshot}.
}
\label{fig:bt_censat}
\end{figure}
%-----------------------------------------------------------------------------------------------------
